Quantum H theorem, variational entropy, and relaxation processes

### 2021-01-22 tex/disscussion.tex
Rewording: sec. 6 (disc. and concl).

\color{black}
\section{Discussions and remarks}\label{sec:disscussion}

@@

The specific forms of the functions $g'_M$ and $g_{nM}$ for classical and quantum systems will be the
subject of our future work. However, some of their properties can be foreseen, \textit{e.g.} they must
be consistent with the system relaxation times, they must be consistent with the mechanisms of 
energy transfer between open systems, there must be related to the cell thermodynamic variables, etc.

@@

In conclusion, we have proposed a framework for studying classical and quantum out-of-equilibrium
diluted gases, whose initial
state is inhomogeneous in nature. This framework consists, briefly, of dividing the system into
a set of small cells, and assuming that each cell is in equilibrium. Subsequently,
$H$-functionals, $\mathcal H$, can be defined
such that the time evolution towards equilibrium resembles the Boltzmann $H$-theorem: $d\mathcal H/dt\leq0$,
and the equality condition is satisfied when the system attains the global equlibrium.
The distribution functions,
stemming from the $\mathcal H$ functionals, at any time, $t$, are the respective equilibrium distribution functions,
which can be Maxwell-Boltzmann, Fermi-Dirac or Bose-Einstein, depending on the classical or quantum nature
of the gas under consideration. Furthermore,
the distribution function of each cell evolves so as to match the global distribution function in equilibrium.
A proof of the relation $d\mathcal H/dt\leq0$ is performed through the variational method,
the functional form of the distribution
functions, in terms of a monotonic function $g$, are defined, and some of the properties of $g$ are discussed.
The functionals $\mathcal H$ can be associated with the entropy of the system under consideration, hence we
refer to this entropy as the variational entropy of the system. Finally,
a potential application of this framework is outlined briefly.

>>>>

\section{Discussion and conclusion}\label{sec:disscussion}

@@

The specific forms of the monotonic functions $g'_M$ and $g_{nM}$ for classical and quantum systems will be the
subject of our future work. However, some of their properties can be foreseen, \textit{e.g.} they must
be consistent with the system relaxation times, they must be consistent with the mechanisms of 
energy transfer between open systems, they must be related to the cell thermodynamic variables, etc.

@@

In conclusion, we have proposed a framework for studying classical and quantum out-of-equilibrium
diluted gases, whose initial
state is inhomogeneous in nature. This framework consists, briefly, of dividing the system into
a set of small cells, and assuming that each cell is in equilibrium. Subsequently,
$H$-functionals, $\mathcal H$, can be defined
such that the time evolution towards equilibrium resembles the Boltzmann $H$-theorem: $d\mathcal H/dt\leq0$,
and the equality condition is satisfied when the system attains the global equlibrium.
The distribution functions,
stemming from the $\mathcal H$ functionals, at any time, $t$, are the respective equilibrium distribution functions,
which can be Maxwell-Boltzmann, Fermi-Dirac or Bose-Einstein, depending on the classical or quantum nature
of the gas under consideration. Furthermore,
the distribution function of each cell evolves so as to match the global distribution function in equilibrium.
A proof of the relation $d\mathcal H/dt\leq0$ is performed through the variational method,
the functional form of the distribution
functions, in terms of a monotonic functions $g$, are defined, and some of the properties of $g$ are discussed.
The functionals $\mathcal H$ can be associated with the entropy of the system under consideration, hence we
refer to this entropy as the variational entropy of the system. Finally,
a potential application of this framework is outlined briefly.

### 2021-01-22 tex/qccorrespondence.tex
Rewording: sec. 6 (disc. and concl).

We start by arguing that, in the classic limit, $f_{nM}\approx0$ for the following reasons.
Systems at very low temperatures, where the quantum effects cannot be ignored, are obviously excluded
from the current analysis. For systems at sufficently large temperature, the particles
occupy almost exclusively high-energy levels. 
Furthermore, the energy spectrum approaches to be a continuum, as is expected by taking the limit
$\hbar\to0$, and the number of particles per level are very close to zero.

>>>>

We start by arguing that, in equilibrium, it is straightforward to proof that
Eq.~(\ref{eq:qfneqtwo}) collapses
into Eq.~(\ref{eq:fbardef}) by taking the limit wherein the 
degeneration parameter $\xi\equiv\exp\big(-(\epsilon-\mu)/(k_BT)\big) \gg 1$.
Alternatively, a more general approach to show the quantum classical correspondence
consists of analyzing the collapse from the quantum to the classical $H$-functionals,
within the appropiated limit.
For the case treated here, this limit is $f_{nM} \approx 0$ because of the following reasons.
Systems at very low temperatures, where the quantum effects cannot be ignored, are obviously excluded
from the current analysis. For systems at sufficently large temperature, the particles
occupy almost exclusively high-energy levels. 
Furthermore, the energy spectrum approaches to be a continuum, as is expected by taking the limit
$\hbar\to0$, and the number of particles per level are very close to zero.

### 2021-01-22 tex/quantum.tex
se agregaron los ultimos cambios, se unió la sección de discusión con conclusiones. La sección de relajación no se ha modificado hasta haber discutido lo que propone el Dr. Carrillo.
Partial commit: rewording remarksanddiscussion
Rewording: sec. 6 (disc. and concl).

%============================================================
\subsubsection{Properties of $\mathcal{H}$ for systems in equilibrium}
%============================================================
If the system is in equilibrium (we will show that it will also be valid at $t\to\infty$), the temperature becomes homogeneous throughout
the complete system. Also, the local number of particles, the local
energy, and the Lagrange multipliers do not depend on the cell number and they should be homogeneous,
this is 
%
\begin{subequations}\label{eq:qEqRestrictions}
\begin{equation}
   {\mathcal{N}}_M(t\to\infty)\equiv \bar{\mathcal{N}}=N/K,
\end{equation}
%
%
\begin{equation}
	{\mathcal{E}}_M(t\to\infty)\equiv \bar{\mathcal{E}}=E/K,
\end{equation}
\end{subequations}
%
%
\begin{subequations}\label{eq:qEqAlphaBeta}
\begin{equation}
	\alpha_M=\bar\alpha,\quad\forall M,
\end{equation}
%
and
\begin{equation}
	\beta_M=\bar\beta, \quad\forall M.
\end{equation}
\end{subequations}
%
Substituting Eqs.~(\ref{eq:qEqAlphaBeta}) into Eq.~(\ref{eq:qfMn}) yields the distribution
function of each cell in equilibrium to be
%
\begin{equation}\label{eq:qfneq}
    \bar f_{Mn}=\bar f_n =\frac{1}{\exp\big(-\bar\alpha-\bar\beta \epsilon_n\big)\pm 1},\quad\forall M.
\end{equation}
%
By means of the above equation, we can recover the distribution function and the entropy of a diluted quantum gas in equilibrium as follows. Setting $\bar\alpha=\mu/kT$ and $\bar\beta=-1/kT$,
and substituting them in Eq.~(\ref{eq:qfneq}), it renders the Fermi-Dirac and Bose-Einstein
distribution functions:
%
\begin{equation}
    \bar{f}_{n}=\frac{1}{\exp\big(\frac{{\epsilon_n}-\mu}{kT}\big)\pm 1},
\end{equation}
%
and substituting Eq.~(\ref{eq:qfneq}) into the negative of Eq.(\ref{eq:qHdef}), 
the entropy of a quantum ideal gas is obtained
%
\begin{eqnarray}\label{eq:varEntropyDef}
      S&=&\sum_{M=1}^{K}
        \sum_n  \left[
        		\left(\frac{1}{\exp\big(-\bar{\alpha}-\bar{\beta}\epsilon_{n}\big)\pm 1} \right)
           \commr{\ln \left(\frac{1}{\exp\big(-\bar{\alpha}-\bar{\beta}\epsilon_{n}\big)\pm 1} \right)}
         \right]\nonumber \\
      &&\quad\pm  \ln \left[\prod_{M=1}^{K} \prod_{n}
         \left(1 \mp \frac{1}{\exp\big(-\bar{\alpha}-\bar{\beta}\epsilon_{n}\big)\pm 1} \right)
         \right] \delta V_M.
\end{eqnarray}
%
This quantity is what we refer to as ``variational entropy'',
and this name reflects the fact that it was obtained \textit{via}
the variational method. We will use it at the end of this article to analyze relaxation processes.

@@

%\begin{proof}
To proof the quantum $H$-theorem, we start by taking the time-derivative of Eq. (\ref{eq:qHdef}):
%
\begin{equation}\label{eq:deltaH}
   \frac{d \mathcal{H} (t)}{dt}= \sum_n \sum_{M=1}^{K} \dot{f}_{nM}(t)\ln \left[ \frac{f_{nM}(t)}{1\mp f_{nM}(t)} \right] \de V_M.
\end{equation}
%
Subsequently, substitute Eq.~(\ref{eq:qFirstOrd}) in the above equation to obtain:
%
\begin{eqnarray}\label{eq:cambioH1}
    \frac{d\mathcal{H} (t)}{dt}&=&
      \sum_n \sum_{M=1}^{K} \bar{f}_{n}\ln \left[
        \frac{\bar{f}_{n}(1+g_{nM})}{1\mp \bar{f}_{n} (1+ g_{nM})}
      \right]\dot{g}_{nM} \delta V_M \nonumber \\
    &=&\sum_n \sum_{M=1}^{K} \bar{f}_n \left \{
      \ln [\bar{f}_n+\bar{f}_n g_{nM}]\dot{g}_{nM}
      -\ln [1\mp\bar{f}_n\mp\bar{f}_n g_{nM}]\dot{g}_{nM}
    \right \}\delta V_M.
\end{eqnarray}
%
By representing the logarithmic terms, corresponding to
Fermi and Bose gases, through a Taylor series around $\bar f_n g_{nM}=0$, and
substituting this into Eq.~(\ref{eq:cambioH1}), one obtains:
%
\begin{subequations}\label{eq:qlnApprox}
\begin{equation}\label{eq:qlnApproxFermions}
	\ln[1\mp\bar{f}_n\mp\bar{f}_n g_{nM}]
    	\approx \ln[1\mp\bar{f}_n]\mp\frac{\bar{f}_n}{1\mp\bar{f}_{n}} g_{nM}
\end{equation}
%
and
%
\begin{equation}\label{eq:qlnApproxBosons}
    \ln [\bar{f}_n+\bar{f}_n g_{nM}] \approx \ln [\bar{f}_n]+ g_{nM},
\end{equation}
\end{subequations}
%
respectively. Eq.~(\ref{eq:qlnApproxFermions}) is valid because, for non-extremely degenerated Fermi gases, 
$1-\bar{f}_n \gg \bar{f}_n|g_{nM}|$, and Eq.~(\ref{eq:qlnApproxBosons}) is fulfilled because,
for Boson gases,
$1+\bar{f}_n \gg \bar{f}_n |g_{nM}|$ when $\bar{f}_n \gg \bar{f}_n |g_{nM}|$.

Combining Eqs. (\ref{eq:cambioH1}) and (\ref{eq:qlnApprox}) yields:
%
\begin{eqnarray}\label{eq:cambioH2}
    \frac{d\mathcal{H}}{dt}&=&\sum_n \sum_{M=1}^{K} \bar{f}_n\left \{ (\ln \bar{f}_n+ g_{nM})\dot{g}_{nM}\right\} \de V_M \delta \epsilon_n \nonumber \\
    &&-\sum_{n}\sum_{M=1}^{K}\bar f_n\left\{ \left( \ln[1\mp\bar{f}_n]\mp \left[\frac{\bar{f}_n}{1\mp\bar{f}_n} \right] g_{nM}\right)\dot{g}_{nM} \right \}\de V_M,
\end{eqnarray}
%
and substituting Eq.~(\ref{eq:qfneq}) into Eq.~(\ref{eq:cambioH2}):
%
\begin{equation}\label{eq:cambioH3}
    \frac{d\mathcal{H}}{dt}=
       \sum_n \sum_{M=1}^{K} \bar{f}_n\left \{
          (\bar{\alpha}+\bar{\beta}{\epsilon}_n)\dot{g}_{nM}
           +g_{nM}\left(1\pm e^{\bar{\alpha}+\bar{\beta}{\epsilon}_n}\right)\dot{g}_{nM}
       \right \} \de V_M.
\end{equation}
%

@@

Substitute the previous expression in Eq.~(\ref{eq:cambioH3}) to obtain:
%
\begin{equation}\label{eq:cambioH4}
   \frac{d\mathcal{H}}{dt}=\sum_n e^{\bar{\alpha}+\bar{\beta}\epsilon_n}
   \sum_M  g_{nM}\dot{g}_{nM} \delta V_M \leq 0.
\end{equation}
%\end{proof}
Thus, the time evolution of the system occurs in such a way that 
$\frac{d\mathcal{H}}{dt}$ is a decreasing function toward zero.
%

@@

Succinctly, considering a diluted quantum gas contained in a vessel of volume $V$ (divided into
$K$ small cells), with total energy $E$ and $N$ quantum free particles, which initially
is out of equilibrium ---but still the local equilibrium hypothesis is valid---,
the functional
%
\begin{eqnarray}\label{eq:qHfunctTheo}
    \mathcal{H} (t)&=&\sum_{M=1}^{K} \sum_{n} \bigg[ f_{Mn}(t) \ln f_{Mn}(t)\pm \Big(1 \mp f_{Mn}(t)) \ln (1 \mp f_{Mn}(t)\Big) \Big]   \delta V_M,
\end{eqnarray}
%
where $f_{Mn}$ is the $M$-th cell distribution function,
evolves in time such that $d\mathcal{H}/dt\leq0$, and the equality 
condition is attained when the system reaches
the equilibrium state. In Eq.~(\ref{eq:qHfunctTheo}), and for a Fermi (Bose)
gas, $f_{Mn}$ corresponds to the Fermi-Dirac (Bose-Einstein) distrubution function for each
cell and locally each cell is in equilibrium, although the complete system may be non-homogeneous,
and is caracterized by the respective $f_{Mn}$, number of particles $\mathcal{N}_M$, energy $\mathcal{E}_M$,
temperature $T_M$, and Legendre Multipliers $\alpha_M$ and $\beta_M$.

>>>>

%============================================================
\subsubsection{Properties of $\mathcal{H}$ for systems in equilibrium}
%============================================================
If the system is in equilibrium (we will show that it will also be valid at $t\to\infty$), the temperature becomes homogeneous throughout
the complete system. Also, the local number of particles, the local
energy, and the Lagrange multipliers do not depend on the cell number and they should be homogeneous,
this is 
%
\begin{subequations}\label{eq:qEqRestrictions}
\begin{equation}
   {\mathcal{N}}_M(t\to\infty)\equiv \bar{\mathcal{N}}=N/K,
\end{equation}
%
%
\begin{equation}
	{\mathcal{E}}_M(t\to\infty)\equiv \bar{\mathcal{E}}=E/K,
\end{equation}
\end{subequations}
%
%
\begin{subequations}\label{eq:qEqAlphaBeta}
\begin{equation}
	\alpha_M=\bar\alpha,\quad\forall M,
\end{equation}
%
and
\begin{equation}
	\beta_M=\bar\beta, \quad\forall M.
\end{equation}
\end{subequations}
%
Substituting Eqs.~(\ref{eq:qEqAlphaBeta}) into Eq.~(\ref{eq:qfMn}) yields the distribution
function of each cell in equilibrium to be
%
\begin{equation}\label{eq:qfneq}
    \bar f_{Mn}=\bar f_n =\frac{1}{\exp\big(-\bar\alpha-\bar\beta \epsilon_n\big)\pm 1},\quad\forall M.
\end{equation}
%
By means of the above equation, we can recover the distribution function and the entropy of a diluted quantum gas in equilibrium as follows. Setting $\bar\alpha=\mu/kT$ and $\bar\beta=-1/kT$,
and substituting them in Eq.~(\ref{eq:qfneq}), it renders the Fermi-Dirac and Bose-Einstein
distribution functions:
%
\begin{equation}\label{eq:qfneqtwo}
    \bar{f}_{n}=\frac{1}{\exp\big(\frac{{\epsilon_n}-\mu}{kT}\big)\pm 1},
\end{equation}
%
and substituting Eq.~(\ref{eq:qfneq}) into the negative of Eq.(\ref{eq:qHdef}), 
the entropy of a quantum ideal gas is obtained
%
\begin{eqnarray}\label{eq:varEntropyDef}
      S&=&\sum_{M=1}^{K}
        \sum_n  \left[
        		\left(\frac{1}{\exp\big(-\bar{\alpha}-\bar{\beta}\epsilon_{n}\big)\pm 1} \right)
           \commr{\ln \left(\frac{1}{\exp\big(-\bar{\alpha}-\bar{\beta}\epsilon_{n}\big)\pm 1} \right)}
         \right]\nonumber \\
      &&\quad\pm  \ln \left[\prod_{M=1}^{K} \prod_{n}
         \left(1 \mp \frac{1}{\exp\big(-\bar{\alpha}-\bar{\beta}\epsilon_{n}\big)\pm 1} \right)
         \right] \delta V_M.
\end{eqnarray}
%
This quantity is what we refer to as ``variational entropy'',
and this name reflects the fact that it was obtained \textit{via}
the variational method. {\color{blue} We will use it at the end of this article to analyze relaxation processes.}

@@

%\begin{proof}
To proof the quantum $H$-theorem, start by taking the time-derivative of Eq. (\ref{eq:qHdef}):
%
\begin{equation}\label{eq:deltaH}
   \frac{d \mathcal{H} (t)}{dt}= \sum_n \sum_{M=1}^{K} \dot{f}_{nM}(t)\ln \left[ \frac{f_{nM}(t)}{1\mp f_{nM}(t)} \right] \de V_M.
\end{equation}
%
Subsequently, substitute Eq.~(\ref{eq:qFirstOrd}) in the above equation to obtain:
%
\begin{eqnarray}\label{eq:cambioH1}
    \frac{d\mathcal{H} (t)}{dt}&=&
      \sum_n \sum_{M=1}^{K} \bar{f}_{n}\ln \left[
        \frac{\bar{f}_{n}(1+g_{nM})}{1\mp \bar{f}_{n} (1+ g_{nM})}
      \right]\dot{g}_{nM} \delta V_M \nonumber \\
    &=&\sum_n \sum_{M=1}^{K} \bar{f}_n \left \{
      \ln [\bar{f}_n+\bar{f}_n g_{nM}]\dot{g}_{nM}
      -\ln [1\mp\bar{f}_n\mp\bar{f}_n g_{nM}]\dot{g}_{nM}
    \right \}\delta V_M.
\end{eqnarray}
%
Approximate the logarithmic terms, corresponding to
Fermi and Bose gases, through a Taylor series around $\bar f_n g_{nM}=0$:
%
\begin{subequations}\label{eq:qlnApprox}
\begin{equation}\label{eq:qlnApproxFermions}
	\ln[1\mp\bar{f}_n\mp\bar{f}_n g_{nM}]
    	\approx \ln[1\mp\bar{f}_n]\mp\frac{\bar{f}_n}{1\mp\bar{f}_{n}} g_{nM}
\end{equation}
%
and
%
\begin{equation}\label{eq:qlnApproxBosons}
    \ln [\bar{f}_n+\bar{f}_n g_{nM}] \approx \ln [\bar{f}_n]+ g_{nM},
\end{equation}
\end{subequations}
%
respectively. Eq.~(\ref{eq:qlnApproxFermions}) is valid because, for non-extremely degenerated Fermi gases, 
$1-\bar{f}_n \gg \bar{f}_n|g_{nM}|$, and Eq.~(\ref{eq:qlnApproxBosons}) is fulfilled because,
for Boson gases,
$1+\bar{f}_n \gg \bar{f}_n |g_{nM}|$ when $\bar{f}_n \gg \bar{f}_n |g_{nM}|$.

Combine Eqs. (\ref{eq:cambioH1}) and (\ref{eq:qlnApprox}):
%
\begin{eqnarray}\label{eq:cambioH2}
    \frac{d\mathcal{H}}{dt}&=&\sum_n \sum_{M=1}^{K} \bar{f}_n\left \{ (\ln \bar{f}_n+ g_{nM})\dot{g}_{nM}\right\} \de V_M \delta \epsilon_n \nonumber \\
    &&-\sum_{n}\sum_{M=1}^{K}\bar f_n\left\{ \left( \ln[1\mp\bar{f}_n]\mp \left[\frac{\bar{f}_n}{1\mp\bar{f}_n} \right] g_{nM}\right)\dot{g}_{nM} \right \}\de V_M,
\end{eqnarray}
%
and substitute Eq.~(\ref{eq:qfneq}) into Eq.~(\ref{eq:cambioH2}):
%
\begin{equation}\label{eq:cambioH3}
    \frac{d\mathcal{H}}{dt}=
       \sum_n \sum_{M=1}^{K} \bar{f}_n\left \{
          (\bar{\alpha}+\bar{\beta}{\epsilon}_n)\dot{g}_{nM}
           +g_{nM}\left(1\pm e^{\bar{\alpha}+\bar{\beta}{\epsilon}_n}\right)\dot{g}_{nM}
       \right \} \de V_M.
\end{equation}
%

@@

Substitute the previous expression in Eq.~(\ref{eq:cambioH3}) to obtain:
%
\begin{equation}\label{eq:cambioH4}
   \frac{d\mathcal{H}}{dt}=\sum_n e^{\bar{\alpha}+\bar{\beta}\epsilon_n}
   \sum_M  g_{nM}\dot{g}_{nM} \delta V_M \leq 0. \qquad\qquad\textrm{QED.}
\end{equation}
%\end{proof}
%

@@

Succinctly, considering a diluted quantum gas contained in a vessel of volume $V$ (divided into
$K$ small cells), with total energy $E$ and $N$ quantum free particles, which initially
is out of equilibrium ---but in such a manner that the local equilibrium hypothesis is valid---,
the functional
%
\begin{eqnarray}\label{eq:qHfunctTheo}
    \mathcal{H} (t)&=&\sum_{M=1}^{K} \sum_{n} \bigg[ f_{Mn}(t) \ln f_{Mn}(t)\pm \Big(1 \mp f_{Mn}(t)) \ln (1 \mp f_{Mn}(t)\Big) \Big]   \delta V_M,
\end{eqnarray}
%
where $f_{Mn}$ is the $M$-th cell distribution function,
evolves in time such that $d\mathcal{H}/dt\leq0$, and the equality 
condition is attained when the system reaches
the equilibrium state. In Eq.~(\ref{eq:qHfunctTheo}), and for a Fermi (Bose)
gas, $f_{Mn}$ corresponds to the Fermi-Dirac (Bose-Einstein) distrubution function for each cell.
Locally, each cell is in equilibrium, although the complete system may be non-homogeneous,
and is caracterized by the respective $f_{Mn}$, number of particles $\mathcal{N}_M$, energy $\mathcal{E}_M$,
temperature $T_M$, and Legendre Multipliers $\alpha_M$ and $\beta_M$.
